---
title: "fsbrain: Visualizing Brain Volume Data From Structural Neuroimaging in GNU R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Brain volume visualization with fsbrain}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

In this document, we demonstrate some high-level functions for loading and visualizing brain volume data. While this package has been developed with  [FreeSurfer](https://surfer.nmr.mgh.harvard.edu) in mind, many of the visualization functions work with 3D matrices and do not make any assumptions on where the data comes from.

Please note that the volume visualization functions presented here are a very recent addition to *fsbrain*, and that they are still in a very early development stage. I am very grateful for any bug reports or comments, please see the [project website](https://github.com/dfsp-spirit/fsbrain).


### Example data used in this document

We will use example data that comes with fsbrain throughout this manual. Feel free to replace the *subjects_dir*, *subjects_list* and *subject_id* with data from your study. 

```{r, eval = FALSE}
library("fsbrain");
fsbrain::download_optional_data();
subjects_dir = fsbrain::get_optional_data_filepath("subjects_dir");
subject_id = 'subject1';       # for function which use one subject only
```

# Loading volume data

If all you want is the raw voxel data, you could load a FreeSurfer volume in MGH or MGZ format like this:


```{r, eval = FALSE}
brain_data = subject.volume(subjects_dir, subject_id, "brain");
```


In many cases, you will need the header information as well, e.g., to be able to compute RAS coordinates for the voxels. If that's the case, load the volume with header:

```{r, eval = FALSE}
brain = subject.volume(subjects_dir, subject_id, "brain", with_header = TRUE);
brain_data = brain$data;
```


# Volume visualization


