---
title: "The fsbrain FAQ"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fsbrain frequently asked questions}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

## Input data

### **Q**: What kind of input data do I need for fsbrain?

The *fsbrain* software is designed to be used with the output of [FreeSurfer](http://freesurfer.net/).  Running *recon-all* on your T1w MRI scan results in a directory structure full of different files and file types for each subject. The *fsbrain* library uses knowledge on this directory layout to load the proper data.

However, fsbrain is not limited to FreeSurfer output, see below.

### **Q**: Which file formats are supported? 

The *fsbrain* library uses [freesurferformats](https://github.com/dfsp-spirit/freesurferformats) to load a variety of neuroimaging file formats, including data exchange formats used by other brain imaging software. See the [freesurferformats website](https://github.com/dfsp-spirit/freesurferformats) for the full list.

### **Q**: I want to load a single file that is *somewhere* on my harddisk, i.e., not within a standard recon-all output directory structure. How can I load it?

You can use [freesurferformats](https://github.com/dfsp-spirit/freesurferformats) directly to load the data, then pass it to fsbrain. See the next question for an example.

### **Q**: Can I use fsbrain to visualize data from SPM12 / CAT12?

Yes, the [computational anatomy toolbox (CAT12)](http://www.neuro.uni-jena.de/cat/) for [SPM](https://www.fil.ion.ucl.ac.uk/spm/) writes surfaces in GIFTI format and the morphometry data in curv format, both formats are supported by *fsbrain*. After running CAT12 surface measure computation on your subject `subject1`, you should have the following files in the *surf/* subdir:

* lh.central.subject1.gii
* lh.gyrification.subject1

Try the following to visualize the gyrification data for the left hemisphere in *fsbrain*:

```{r, eval = FALSE}
lh_surf = freesurferformats::read_nisurface('~/data/subject1_spm12/surf/lh.central.subject1.gii');
lh_gyrification = freesurferformats::read.fs.curv('~/data/subject1_spm12/surf/lh.gyrification.subject1');
vis.data.on.subject('~/data/', 'subject1_spm12', lh_gyrification, NULL, surface=lh_surf);
```

## Visualization


### How can I increase the output image resolution?

To increase the output resolution, you need to increase the size of the *rgl* rendering device. To do this globally, before you call any fsbrain rendering function:

```{r, eval = FALSE}
library('rgl');
r3dDefaults$windowRect=c(20, 20, 1800, 1200);
```

This instructs *rgl* to create new rendering devices at screen position 20, 20 (close to the upper left corner) with a size of 1800x1200 pixels.

Alternatively, you can control the size when calling an *fsbrain* visualization function by passing the same information in the optional `rgloptions` parameter, like this:

```{r, eval = FALSE}
rglo = list('windowRect'=c(20, 20, 1800, 1200));
vis.subject.morph.native('~/mysubjects_dir', 'subject1', 'thickness', rgloptions=rglo);
```

###  How can I save high quality figures to PNG images?

Note that fsbrain renders images, which means the output is pixel-based (i.e., bitmap as opposed to vector graphics). To get high quality output, you need to increase the size of the *rgl* rendering device, as explained in the last question.

To save the plot to a file in PNG format, you can use an rglaction:

```{r, eval = FALSE}
rgla = list('snapshot_png'='~/subject1_thickness.png');
vis.subject.morph.native('~/mysubjects_dir', 'subject1', 'thickness', rglactions=rgla);
```

This opens the plot in a window as usual and also saves it in PNG format to the file *subject1_thickness.png* in your home directory.

