---
title: "fsbrain visualization examples"
author: "Tim Schäfer"
date: "April 23, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rgl::setupKnitr()
```


## Example data

First we specify a subjects dir and the subject we want to use. Feel free to replace this with your data.

```{r}
library('fsbrain')
sjd = '~/data/subject1_only'
sj = 'subject1'
```

Now increase the figure size.

```{r}
library('rgl')
r3dDefaults$windowRect=c(20, 20, 900, 900);
```

## Visualizing native space morphometry data: cortical thickness

Let's start with a basic example and plot cortical thickness on the surface of our subject:

```{r rgl=TRUE, dev='png'}
vis.subject.morph.native(sjd, sj, 'thickness')
```

This loaded the morphometry data files `surf/lh.thickness` and `surf/rh.thickness` of the subject, applied a colormap to the values, and visualized the results on the white surface, i.e., the brain surface meshes in the files `surf/lh.white` and `surf/rh.white`.

## Changing the view and masking the medial wall

The default view above shows the brain from four different angles in four tiles and is called `t4` view. Maybe you prefer a different view? In addition, let's mask the medial wall (the large blue parts in the lower two tiles).

```{r rgl=TRUE, dev='png'}
vis.subject.morph.native(sjd, sj, 'thickness', views='t9', cortex_only = TRUE)
```

The mask for the medial wall was created by setting the data of all vertices which are not part of the cortex to `NA`, based on the cortex definition in the label files `label/lh.cortex.label` and `label/rh.cortex.label` of the subject.

## Visualizing data with outliers, tails, or other non-normal data

Next, we will try a third view, add a colorbar, use the pial surface, and plot something different. Let's try mean curvature.

```{r rgl=TRUE, dev='png'}
vis.subject.morph.native(sjd, sj, 'curv', views='si', cortex_only = TRUE, draw_colorbar = 'vertical', surface = 'pial')
```

That does not look very exciting. Why? The curvature data has long tails and is far from Normal:

```{r}
hist(subject.morph.native(sjd, sj, 'curv', hemi = 'both'))
```

There are various solutions to this problem, but for plotting, the easiest one is to apply some function that removes outliers or performs some other sort of data transformation before plotting. You can pass any transformation function you want using *rglactions*, in the next example we clip the data.

The `si` view is single interactive view, and you can rotate the brain with your mouse in an R session. That is not possible in this markdown document though, so better try it in R. We will use another view here and switch back to the white surface.

```{r rgl=TRUE, dev='png'}
vis.subject.morph.native(sjd, sj, 'curv', views='sd_medial_lh', cortex_only = TRUE, draw_colorbar = 'horizontal',  rglactions=list('trans_fun'=clip.data))
```

That looks more suitable.


## Standard space data


In group studies, you typically have a single standard space that data gets mapped to for comparison. For FreeSurfer surface data, typically the *fsaverage* subject is used, but you could create and use your own template, of course. Let's have a look at some standard space data mapped to fsaverage. We will use sulcal depths in this example.

```{r rgl=TRUE, dev='png'}
vis.subject.morph.standard(sjd, sj, 'sulc', cortex_only = TRUE, fwhm="5");
```


This loaded the morphometry data files `surf/?h.curv.fwhm5.fsaverage.mgh` of the subject, applied a colormap to the values, and visualized the results on the white surface of fsaverage (also using the fsaverage cortex labels). The function will do its best to find fsaverage automatically, but if it fails or you want to use a different template subject, have a look at the optional parameters `template_subject` and `template_subjects_dir`.

When data is mapped to standard space, they will get smoothed and different versions with a number of kernel settings will be available after running FreeSurfer's `recon-all` with the `qcache` option. That is what the `fwhm` parameter is about. Try different values depending on what files you have, start with "5", "10", or "15".


## Adjusting the colormap

The exact colormap you use is a matter of taste, but you should maybe chose a type of colormap that is suitable for your data: *sequential* like `heat` for morphometry data, *diverging* like `cm.colors` for positive and negative clusters, or *qualitative* like `Set3` from the `RColorbrewer` package for atlas regions. The default colormap used by fsbrain is a sequential multi-hue palette (`squash::rainbow2`).

A great and popular sequential palette is `viridis` from the `viridis` package. Here we used the `makecmap_options` parameter to use it:

```{r rgl=TRUE, dev='png'}
vis.subject.morph.standard(sjd, sj, 'sulc', cortex_only = TRUE, fwhm="15", makecmap_options = list('colFn'=viridis::viridis), draw_colorbar = T, views='sd_dorsal');
```


Try some different colormaps and see which one you like best. If you want to use interpolated colormaps with many colors based on the `RColorBrewer` palettes (which have too few colors for morphometry data in most cases), try something like this:

```{r rgl=TRUE, dev='png'}
manyblues = colorRampPalette(RColorBrewer::brewer.pal(9, name="Blues"));
vis.subject.morph.standard(sjd, sj, 'area', makecmap_options = list('colFn'=manyblues), draw_colorbar = T, views='sd_dorsal');
```




